\bf{Neural Ordinary Differential Equations}

### 2025-03-20 paper.tex
completed the report finally

In most engineering systems, we only have access to noisy measurements of 
input-output data, typically originating from high-dimensional, nonlinear dynamic 
systems. While such systems can be approximated using finite impulse response (FIR) 
or autoregressive exogenous (ARX) models, these approaches often fail to respect 
fundamental physical laws. One way to address this limitation is by penalizing 
violations of these laws—a key idea behind physics-informed neural networks (PINNs), 
as introduced by \cite{raissi:perdikaris:karniadakis:2019}.

However, it is important to view PINNs not merely as data-driven models but 
rather as a solution strategy for PDEs and ODEs. Fundamentally, they operate 
as \textit{methods of weighted residuals}—specifically, collocation schemes—where 
the residual is minimized at selected collocation points. While PINNs can be 
used to learn certain model parameters, their results can be difficult to interpret 
without a fundamental understanding of the underlying physics.

@@

Here we give a quick overview of both the forward and backward sensitivity/adjoint 
equations for ODEs. Consider following non-linear state-space model:
\begin{align}
    \dot{x} &= f(x, u; \theta) \qquad x(0) = g(x_0; \theta)\notag 
    \label{eq:ssode}
\end{align}
We assume full state measurement (i.e. $y=x$). Consider a least-squares objective
to measure the fit to the data ($\tilde{x}$):
\begin{align}
    \min\limits_{\theta} V(x) = \norm{\tilde{x} - x}^2 
\end{align}
To propagate the gradient we need to solve the sensitivity equations:
\begin{align}
    \dfrac{\partial V}{\partial \theta^T} &= -2 (\tilde{x} - x)^T \dfrac{\partial x}{\partial \theta^T} \notag \\ 
    S &= \dfrac{\partial x}{\partial \theta^T}
    \label{eq:sens}
\end{align}
where $S$ is the \textit{sensitivity} matrix of the model solutions with parameters. We differentiate $S$ to obtain:
\begin{align}
    \dfrac{d S}{dt} &= \dfrac{d}{dt} \left(\dfrac{\partial x}{\partial \theta^T}\right) =
    \dfrac{\partial}{\partial\theta^T} \left(\dfrac{d x}{dt}\right) = 
    \dfrac{\partial }{\partial \theta^T} f
\end{align}
Using chain rule on $f$, we get following matrix equation:
\begin{align}
    \dfrac{dS}{dt} = \dfrac{\partial f}{\partial x^T}S + \dfrac{\partial f}{\partial \theta^T} \qquad  S(0) = \dfrac{\partial g}{\partial \theta^T}
\end{align}
Thus, we can construct following augmented ODE system:
\begin{align}
    \begin{bmatrix} \dot{x} \\ \dot{S} \end{bmatrix} = 
    \begin{bmatrix} f \\ \dfrac{\partial f}{\partial x^T}S + \dfrac{\partial f}{\partial \theta^T} \end{bmatrix}
    \qquad
    \begin{bmatrix} x(0) \\ S(0) \end{bmatrix} =
    \begin{bmatrix} x_0 \\ \dfrac{\partial g}{\partial \theta^T} \end{bmatrix}
    \label{eq:augsensode}
\end{align}
This augmented system can be solved using any ODE solver, allowing gradient 
propagation. However, if $\theta \in \R^{p}$ and $x \in \R^{n}$, we need to 
solve $S \in \R^{n+p}$ equations i.e. the cost increases with $p$. Thus, forward
sensitivity is not an efficient method to propagate gradients. However, we can circumvent
the problem of linear increase in sensitivity equations using backward adjoint equations. This key advantage 
of backward adjoint method contributed to the feasibility of Neural ODEs (NODEs), beyond their interpretation 
as a continuous counterpart to Euler-step based architectures like ResNets. We next show 
a short example on solving the augmented ODE system for forward sensitivity. Consider, 
following kinetics:
\begin{align}
    \dot{c}_A &= -kc_A \notag \\
    c_A(0) &= c_{A0}
\end{align}
Say we want to calculate sensitivities wrt $k$ and $c_{A0}$. With reference to \cref{eq:augsensode},
for $S_1 = \dfrac{\partial c_A}{\partial k}$ and $S_2 = \dfrac{\partial c_A}{\partial c_{A0}}$,
we have following augmented system:
\begin{align}
    \begin{bmatrix} 
        \dot{c}_A \\
        \dot{S_1} \\ 
        \dot{S_2} 
    \end{bmatrix} = 
    \begin{bmatrix}
        -kc_A \\
        -kS_1 - c_A \\
        -kS_2 \end{bmatrix}
    \qquad
    \begin{bmatrix} 
        c_A(0) \\ 
        S_1(0) \\ 
        S_2(0) 
    \end{bmatrix} =
    \begin{bmatrix} 
        c_{A0} \\ 
        0 \\ 
        1
    \end{bmatrix}
\end{align}
If we were learning the parameter $k$ and initial condition $c_{A0}$, we could
show a gradient descent update as follows:
\begin{align}
    \begin{bmatrix} k \\ c_{A0} \end{bmatrix} \leftarrow
    \begin{bmatrix} k \\ c_{A0} \end{bmatrix}
    - \eta 
    \begin{bmatrix} 
        -2 (\hat{c}_A  - c_A)^T S_1 \\ 
        -2 (\hat{c}_A  - c_A)^T S_2
    \end{bmatrix}
\end{align}

@@

In this section we redefine the objective function as an integral over time.
The integral formulation allows us to derive the adjoint equation.
Consider following least-squares objective for \cref{eq:ssode} and the Lagrange functional:

@@

\begin{align}
    \dfrac{d\lambda^T}{dt} &= -\dfrac{\partial g}{\partial x^T} - \lambda^T \dfrac{\partial f}{\partial x^T} \notag \\
    \lambda(T) &= 0
\end{align}
Referring to \cref{eq:adjoint}, we see $\dfrac{\partial V}{\partial \theta^T} = 
\dfrac{\partial \mathcal{L}}{\partial \theta^T}$.
The original NODE paper uses objective function of the form:
\begin{align}
    \min\limits_{\theta}V(x) = \int_0^t \norm{x - \hat{x}}^2 \delta(t) dt
\end{align}
Hence, the adjoint equations for their case are:
\begin{align}
    \dfrac{d\lambda^T}{dt} &= -\lambda^T \dfrac{\partial f}{\partial x^T} \notag \\
    \lambda(t)^T &= - \dfrac{\partial g}{\partial x^T}|_t = -\dfrac{\partial V}{\partial x^T}|_t \notag \\
    \dfrac{\partial V}{\partial \theta^T} &= \lambda^T(0)\dfrac{\partial x}{\partial \theta^T}|_0 + \int_0^t   \lambda^T \dfrac{\partial f}{\partial \theta^T} dt
    \label{eq:adjointnode}
\end{align}
Hence, $\lambda = \dfrac{\partial V}{\partial x}$ as we see from the terminal 
condition in \cref{eq:adjointnode}.
Thus, we need to solve the augmented ODE system once forward 
and once backward in each optimal iteration. We need a forward pass to store
the value of $\dfrac{\partial g}{\partial x^T}|_t$ and $x_t$. Also, note that the number 
of adjoint equations do not increase with parameters (i.e $\lambda \in \R^n$ unlike $S \in \R^{n \times p+1}$). 
We can evaluate the gradient in one go as shown in \cref{eq:adjointnode}. 
Some recent solutions to storing intermediate points involve \textit{checkpointing}
at only few points. However, such a method is not 
an ultimate panacea. Such systems are still vulnerable to the problems commonly faced by
numerical ODE solvers like stiffness, accuracy, etc. One can analyze the stiffness
of the adjoint ODE by simply inspecting the eignevalues of the Jacobian matrix $\dfrac{\partial f}{\partial x^T}$
at different time points. Sometimes we may end up with noisy gradients which may lead to poor convergence.

@@

The NODE paper applies the same motivation to develop continuous normalizing flows.
Interested reader can refer the paper for theorem on instantaneous change of variables.
The key idea is the neural ODE rhs is trained to conserve the probability density 
while maximizing log-likelihood to describe the data. The framework can be described 
as follows:
\begin{align}
    \begin{bmatrix}
        x_0 \\
        \log p(x) - \log p(x_0)
    \end{bmatrix}
    =
    \int_{t}^{0}
    \begin{bmatrix}
        f_{NN}(x, u; \theta) \\
        -\mathrm{Tr} \left(\dfrac{\partial f_{NN}}{\partial x^T}\right)
    \end{bmatrix} dt;
    \qquad
    \begin{bmatrix}
        x(t) \\
        \log p(x) - \log p(x_t)
    \end{bmatrix}   
    =
    \begin{bmatrix}
        x_t \\
        0
    \end{bmatrix}
\end{align}
We assume, $x_0 \sim \mathcal{N}(0, \sigma^2I)$ and thus the objective function is:
\begin{align}
    \min\limits_{\theta} V(x) = -\log p(x)
\end{align}
That's a powerful framework to model distributions (say a human face composed of pixels).
However, we are more interested in using similar generative framework for chemical kinetics. 
Thus, we again use the same example as shown in \cref{eq:atobtoc} but we also model 
how the uncertainty in the initial conditions decreases as we add more measurement points.
The results are shown in \cref{fig:cnf}. The system we solve is:

\begin{align}
    \begin{bmatrix} \dot{c}_A \\ \dot{c}_B \\ \dot{c}_C \\ \dot{\log p} \end{bmatrix}
    = \begin{bmatrix}
        -k_{1}c_A + k_{-1}c_B  \\
        2k_{1}c_A +  2k_{-1}c_B - k_2c_B \\
        k_{2}c_B\\
        -\mathrm{Tr} \left(\dfrac{\partial f_{NN}}{\partial x^T}\right)
    \end{bmatrix};
    &\qquad
    \begin{bmatrix} c_A(0) \\ c_B(0) \\ c_C(0) \\ \log p(0) \end{bmatrix} =
    \begin{bmatrix} 1 \\ 0 \\ 0 \\ \log{\mathcal{N}([c_A, c_B, c_C]^T, \sigma^2I)} \end{bmatrix} \notag \\
    \min\limits_{\theta} V(x) &= -\log p(x) + \norm{c - \tilde{c}}^2
\end{align}
The author would like to draw parallels to the classical Kalman filter that also does the similar job,
but the continuous normalizing flow can be applied to any non-linear dynamical system.
\begin{figure}[h]
    \centering
    \includegraphics[width=1\textwidth, page=3]{ABC_plot.pdf} 
    \caption{Results for chemical kinetics modeled as continuous normalizing
    flow. We start with a high uncertainty in the initial conditions. The spread in 
    probability decreases as we add more measurement points. Readers would like
    to draw similarities to Kalman filter, only in this case we can apply it to 
    any non-linear continuous dynamical system.}
    \label{fig:cnf}
\end{figure}

@@

\section{Conclusions}

>>>>

In most engineering systems, we only have access to noisy measurements of 
input-output data, typically originating from high-dimensional, nonlinear dynamic 
systems. While such systems can be approximated using finite impulse response (FIR) 
or autoregressive exogenous (ARX) models \citep{ljung:1999}, these approaches often fail to respect 
fundamental physical laws. One way to address this limitation is by penalizing 
violations of these laws—a key idea behind physics-informed neural networks (PINNs), 
as introduced by \cite{raissi:perdikaris:karniadakis:2019}.

However, it is important to view PINNs not merely as data-driven models but 
rather as a solution strategy for PDEs and ODEs. Fundamentally, they operate 
as \textit{methods of weighted residuals} \citep{villadsen:stewart:1967}—specifically, collocation schemes—where 
the residual is minimized at selected collocation points. While PINNs can be 
used to learn certain model parameters, their results can be difficult to interpret 
without a fundamental understanding of the underlying physics.

@@

Here we give a quick overview of both the forward and backward sensitivity/adjoint 
equations for ODEs. Consider following non-linear state-space model:
\begin{align}
    \dot{x} &= f(x, u; \theta) \qquad x(0) = g(x_0; \theta)\notag 
    \label{eq:ssode}
\end{align}
We assume full state measurement (i.e. $y=x$). Consider a least-squares objective
to measure the fit to the data ($\tilde{x}$):
\begin{align}
    \min\limits_{\theta} V(x) = \norm{\tilde{x} - x}^2 
\end{align}
We refer \cite{rawlings:ekerdt:2020} for following derivation. To propagate the gradient we need to solve the sensitivity equations:
\begin{align}
    \dfrac{\partial V}{\partial \theta^T} &= -2 (\tilde{x} - x)^T \dfrac{\partial x}{\partial \theta^T} \notag \\ 
    S &= \dfrac{\partial x}{\partial \theta^T}
    \label{eq:sens}
\end{align}
where $S$ is the \textit{sensitivity} matrix of the model solutions with parameters. We differentiate $S$ to obtain:
\begin{align}
    \dfrac{d S}{dt} &= \dfrac{d}{dt} \left(\dfrac{\partial x}{\partial \theta^T}\right) =
    \dfrac{\partial}{\partial\theta^T} \left(\dfrac{d x}{dt}\right) = 
    \dfrac{\partial }{\partial \theta^T} f
\end{align}
Using chain rule on $f$, we get following matrix equation:
\begin{align}
    \dfrac{dS}{dt} = \dfrac{\partial f}{\partial x^T}S + \dfrac{\partial f}{\partial \theta^T} \qquad  S(0) = \dfrac{\partial g}{\partial \theta^T}
\end{align}
Thus, we can construct following augmented ODE system:
\begin{align}
    \begin{bmatrix} \dot{x} \\ \dot{S} \end{bmatrix} = 
    \begin{bmatrix} f \\ \dfrac{\partial f}{\partial x^T}S + \dfrac{\partial f}{\partial \theta^T} \end{bmatrix}
    \qquad
    \begin{bmatrix} x(0) \\ S(0) \end{bmatrix} =
    \begin{bmatrix} x_0 \\ \dfrac{\partial g}{\partial \theta^T} \end{bmatrix}
    \label{eq:augsensode}
\end{align}
This augmented system can be solved using any ODE solver, allowing gradient 
propagation. However, if $\theta \in \R^{p}$ and $x \in \R^{n}$, we need to 
solve $S \in \R^{n+p}$ equations i.e. the cost increases with $p$. Thus, forward
sensitivity is not an efficient method to propagate gradients. However, we can circumvent
the problem of linear increase in sensitivity equations using backward adjoint equations. This key advantage 
of backward adjoint method contributed to the feasibility of Neural ODEs (NODEs), beyond their interpretation 
as a continuous counterpart to Euler-step based architectures like ResNets. We next show 
a short example on solving the augmented ODE system for forward sensitivity. Consider, 
following kinetics:
\begin{align}
    \dot{c}_A &= -kc_A \notag \\
    c_A(0) &= c_{A0}
\end{align}
Say we want to calculate sensitivities wrt $k$ and $c_{A0}$. With reference to \cref{eq:augsensode},
for $S_1 = \dfrac{\partial c_A}{\partial k}$ and $S_2 = \dfrac{\partial c_A}{\partial c_{A0}}$,
we have following augmented system:
\begin{align}
    \begin{bmatrix} 
        \dot{c}_A \\
        \dot{S_1} \\ 
        \dot{S_2} 
    \end{bmatrix} = 
    \begin{bmatrix}
        -kc_A \\
        -kS_1 - c_A \\
        -kS_2 \end{bmatrix}
    \qquad
    \begin{bmatrix} 
        c_A(0) \\ 
        S_1(0) \\ 
        S_2(0) 
    \end{bmatrix} =
    \begin{bmatrix} 
        c_{A0} \\ 
        0 \\ 
        1
    \end{bmatrix}
\end{align}
If we were learning the parameter $k$ and initial condition $c_{A0}$, we could
show a gradient descent update as follows:
\begin{align}
    \begin{bmatrix} k \\ c_{A0} \end{bmatrix} \leftarrow
    \begin{bmatrix} k \\ c_{A0} \end{bmatrix}
    - \eta 
    \begin{bmatrix} 
        -2 (\hat{c}_A  - c_A)^T S_1 \\ 
        -2 (\hat{c}_A  - c_A)^T S_2
    \end{bmatrix}
\end{align}

@@

In this section we redefine the objective function as an integral over time.
The integral formulation allows us to derive the adjoint equation. We refer \cite{sengupta:friston:penny:2014} 
that gives a straightforward proof of the adjoint equation than the one in
\cite{chen:rubanova:bettencourt:duvenaud:2018}.
Consider following least-squares objective for \cref{eq:ssode} and the Lagrange functional:

@@

\begin{align}
    \dfrac{d\lambda^T}{dt} &= -\dfrac{\partial g}{\partial x^T} - \lambda^T \dfrac{\partial f}{\partial x^T} \notag \\
    \lambda(T) &= 0
\end{align}
Referring to \cref{eq:adjoint}, we see $\dfrac{\partial V}{\partial \theta^T} = 
\dfrac{\partial \mathcal{L}}{\partial \theta^T}$.
\cite{chen:rubanova:bettencourt:duvenaud:2018} uses objective function of the form:
\begin{align}
    \min\limits_{\theta}V(x) = \int_0^t \norm{x - \hat{x}}^2 \delta(t) dt
\end{align}
Hence, the adjoint equations for their case are:
\begin{align}
    \dfrac{d\lambda^T}{dt} &= -\lambda^T \dfrac{\partial f}{\partial x^T} \notag \\
    \lambda(t)^T &= - \dfrac{\partial g}{\partial x^T}|_t = -\dfrac{\partial V}{\partial x^T}|_t \notag \\
    \dfrac{\partial V}{\partial \theta^T} &= \lambda^T(0)\dfrac{\partial x}{\partial \theta^T}|_0 + \int_0^t   \lambda^T \dfrac{\partial f}{\partial \theta^T} dt
    \label{eq:adjointnode}
\end{align}
Hence, $\lambda = \dfrac{\partial V}{\partial x}$ as we see from the terminal 
condition in \cref{eq:adjointnode}.
Thus, we need to solve the augmented ODE system once forward 
and once backward in each optimal iteration. We need a forward pass to store
the value of $\dfrac{\partial g}{\partial x^T}|_t$ and $x_t$. Also, note that the number 
of adjoint equations do not increase with parameters (i.e $\lambda \in \R^n$ unlike $S \in \R^{n \times p+1}$). 
We can evaluate the gradient in one go as shown in \cref{eq:adjointnode}. 
However, such method is not 
an ultimate panacea. Such systems are still vulnerable to the problems commonly faced by
numerical ODE solvers like stiffness, accuracy, etc. One can analyze the stiffness
of the adjoint ODE by simply inspecting the eignevalues of the Jacobian matrix $\dfrac{\partial f}{\partial x^T}$
at different time points. Sometimes we may end up with noisy gradients which may lead to poor convergence.
Some recent remedies is solving adjoint equation in multiple time-intervals rather than 
a single long time-interval
and storing intermediate points, often called as \textit{checkpointing}.
 \citep{zhuang:dvornek:li:tatikonda:papademetris:duncan:2020}.

@@

The NODE paper applies the same motivation to develop continuous normalizing flows
\citep{grathwohl:chen:bettencourt:sutskever:duvenaud:2018}.
Interested reader can refer the paper for proof of theorem on instantaneous change of variables
\citep{chen:rubanova:bettencourt:duvenaud:2018}.
The key idea is the neural ODE rhs is trained to conserve the probability density 
while maximizing log-likelihood to describe the data. The framework can be described 
as follows:
\begin{align}
    \begin{bmatrix}
        x_0 \\
        \log p(x) - \log p(x_0)
    \end{bmatrix}
    =
    \int_{t}^{0}
    \begin{bmatrix}
        f_{NN}(x, u; \theta) \\
        -\mathrm{Tr} \left(\dfrac{\partial f_{NN}}{\partial x^T}\right)
    \end{bmatrix} dt;
    \qquad
    \begin{bmatrix}
        x(t) \\
        \log p(x) - \log p(x_t)
    \end{bmatrix}   
    =
    \begin{bmatrix}
        x_t \\
        0
    \end{bmatrix}
\end{align}
We assume, $x_0 \sim \mathcal{N}(0, \sigma^2I)$ and thus the objective function is:
\begin{align}
    \min\limits_{\theta} V(x) = -\log p(x)
\end{align}
That's a powerful framework to model distributions (say a human face composed of pixels).
However, we are more interested in using similar generative framework for chemical kinetics. 
Thus, we again use the same example as shown in \cref{eq:atobtoc} but we also model 
how the uncertainty in the initial conditions decreases as we add more measurement points.
The results are shown in \cref{fig:cnf}. The system we solve is:

\begin{align}
    \begin{bmatrix} \dot{c}_A \\ \dot{c}_B \\ \dot{c}_C \\ \dot{\log p} \end{bmatrix}
    = \begin{bmatrix}
        -k_{1}c_A + k_{-1}c_B  \\
        2k_{1}c_A +  2k_{-1}c_B - k_2c_B \\
        k_{2}c_B\\
        -\mathrm{Tr} \left(\dfrac{\partial f_{NN}}{\partial x^T}\right)
    \end{bmatrix};
    &\qquad
    \begin{bmatrix} c_A(0) \\ c_B(0) \\ c_C(0) \\ \log p(0) \end{bmatrix} =
    \begin{bmatrix} 1 \\ 0 \\ 0 \\ \log{\mathcal{N}([c_A, c_B, c_C]^T, \sigma^2I)} \end{bmatrix} \notag \\
    \min\limits_{\theta} V(x) &= -\log p(x) + \norm{c - \tilde{c}}^2
\end{align}
The author would like to draw parallels to the classical Kalman filter \citep{kalman:1960} that also does the similar job,
but the continuous normalizing flow can be applied to any non-linear dynamical system.
\begin{figure}[h]
    \centering
    \includegraphics[width=1\textwidth, page=3]{ABC_plot.pdf} 
    \caption{Results for chemical kinetics modeled as continuous normalizing
    flow. We start with a high uncertainty in the initial conditions. The spread in 
    probability decreases as we add more measurement points. Readers would like
    to draw similarities to Kalman filter, only in this case we can apply it to 
    any non-linear continuous dynamical system.}
    \label{fig:cnf}
\end{figure}

@@

Well-posedness of an IVP is given by following theorem \citep{ascher:petzold:1998}:
\begin{theorem}
    Let $f(t,x)$ be continuous for all $(t,x)$ in a region $\mathcal{D} = {0 \leq t \leq b, |x| \leq \infty}$. 
    Moreover assume Lipschitz contnuity in $x$, then there exists a constant $L$ such that:
    \begin{align*}
        \norm{f(t,x) - f(t,\bar{x})} \leq L \norm{x-\bar{x}}
    \end{align*}   
\end{theorem}
Thus, for ODEs with slightly perturbed initial conditions, the solution should not
intersect though they can approach the same trajectory. Thus, the solution of an 
ODE can be sandwhiched between two other perturbed solutions. Refer to \cref{fig:robust} for
a simple example that again follows the kinetics in \cref{eq:atobtoc}.
\begin{figure}[h]
    \centering
    \includegraphics[width=0.55\textwidth, page=5]{ABC_plot.pdf} 
    \caption{Robustness analysis of ODEs. We see that the solution $c_{B2}$ 
    is sandwiched between two perturbed solutions $c_{B1}, c_{B3}$, 
    though all approach the same steady-state. Also note that the slightly time-shifted
    solution appears as a perturbed solution}
    \label{fig:robust}
\end{figure}
Thus, following above idea, \cite{yan:du:tan:feng:2019} developed robust NODEs.
The idea is the perturbed solution can be seen as time-shifted solution of the
original ODE sytem. Using uniqueness of solutions, time-invariant NODEs can be regularized by
adding a penalty term to the loss function. The penalty term is given by:
\begin{align*}
    \norm{x(t+T)-x(t)} \leq \norm{\int_t^{t+T} f_{NN}(x, u; \theta) dt}
\end{align*}

\begin{figure}[h]
    \centering
    \includegraphics[width=0.55\textwidth, page=6]{ABC_plot.pdf} 
    \caption{If we train based only on prediction loss, the model can overfit.
    However, if we penalize the jacobian of the loss, we can regularize the model
    to damp the oscillations.}
    \label{fig:jacobnode}
\end{figure}

\cite{finlay:jacobsen:nurbekyan:oberman:2020} recommend to penalize the Jacobian of ODE rhs function
$(\nabla_x f)$ to damp oscillations that happen due to overfitting in neural ODE. A simple example
using polynomial fit is show in \cref{fig:jacobnode}.
\section{Conclusions}
In this review, we explored the mathematical foundations and practical applications 
of Neural Ordinary Differential Equations (NODEs), highlighting their connection 
to classical ODE-based modeling and modern machine learning approaches. We first 
established how NODEs extend traditional discrete architectures such as ResNets 
and RNNs by formulating neural networks as continuous dynamical systems. This 
formulation provides a more principled way to model physical and process-driven 
data while reducing memory constraints through adjoint-based sensitivity analysis.

A key focus of this review was the gradient computation techniques for NODEs. We 
discussed the forward sensitivity method, which requires solving an augmented ODE 
system but suffers from an increase in computational cost with the number of parameters. 
The adjoint sensitivity method, in contrast, offers a more efficient alternative by 
solving a backward ODE, making NODEs feasible for large-scale learning problems. While 
adjoint methods significantly improve efficiency, they are not free from challenges, 
as stiffness and numerical instabilities can still affect convergence.

We also explored continuous normalizing flows (CNFs), where NODEs are utilized to 
construct generative models that explicitly model probability density evolution. 
The application of CNFs to chemical kinetics demonstrated how this framework can 
be extended to physical modeling while drawing conceptual parallels with classical 
filtering techniques such as the Kalman filter. These methods provide powerful tools 
for uncertainty quantification and data-driven discovery in dynamical systems.

A critical aspect of NODE-based learning is the robustness of solutions. We examined 
how NODEs, as solutions to differential equations, inherit the stability properties 
of the underlying ODEs. In particular, well-posedness ensures that small perturbations 
in initial conditions do not lead to solution intersections, reinforcing the interpretability 
and reliability of NODE-based models. This insight was leveraged to propose robust NODEs, 
where regularization techniques enforce time-invariance by penalizing deviations between time-shifted trajectories.

Overall, NODEs present a compelling framework for modeling dynamical systems in a continuous, 
memory-efficient, and theoretically grounded manner.

\section{Data availability and Code}
All the figures in the review have been generated by the author. The code for the
figures can be found at
